---
title: "An Introduction to cosinet"
author: "Lanying Wei"
package: "cosinet"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{An introduction to cosinet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    error = TRUE,
    warning = TRUE,
    message = TRUE,
    comment = "#>",
    crop = NULL,
    out.width = "100%"
)
```

# Introduction

`Cosinet` is an R package that aims to bridge the gap between differential
co-expression network analysis and precision medicine by accurately
quantifying the degree of rewiring of co-expression networks at
the level of individual samples. Differential co-expression network analysis
is a valuable approach for understanding the underlying mechanisms of disease
and informing the development of personalized therapies, but current
approaches typically measure an average differential network across multiple
samples, making it difficult to apply in the field of precision medicine.
Cosinet utilizes gene expression data to determine the degree of similarity
between the gene co-expression patterns of a given sample and reference
conditions within the context of a function-specific differential
co-expression network. To obtain differential co-expression network, 
we provide R function for z-score-based differential co-expression analysis. 
Next, we calculate eigenvector centrality to get a ranked gene list, 
which represents the degree of hubness of each gene in the network.
Based on the ranking, one can further perform gene set enrichment analysis
(GSEA) to get enriched pathways, and use the core enrichment genes of a relevant 
pathway to form a function-specific differential co-expression sub-network. 
Finally, Cosinet scores can be calculated based on the specific 
sub-network. Here we show an example of using Cosinet to quantify the degree of 
rewiring with respect to the estrogen receptor (ER) response network for each 
estrogen receptor-positive (ER+) sample relative to the whole ER+ and estrogen 
receptor-negative (ER–) groups. ER+ breast cancer patients are typically treated 
with endocrine therapy. We demonstrate the quantified Cosinet scores 
significantly associate with survival outcomes in ER+ samples treated with 
endocrine therapy. The results highlight the potential of Cosinet to identify 
important differences in gene co-expression patterns between individual patients 
and to provide valuable insights for personalized treatment decisions.

# Installation
To install the cosinet package, you can use the remotes package:
```{r setup, eval=FALSE}
# install Cosinet
if (!requireNamespace("remotes", quietly=TRUE))
    install.packages("remotes")
remotes::install_github("LanyingWei/cosinet")

```

# Data analysis example

## Prepare data
First we load the library and an exemplar dataset, and preprocess the expression 
data. The exemplar data is a subset of RNA-seq gene expression data from breast 
cancer patients. The full data is available in the Gene Expression Omnibus (GEO) repository under accession number GSE96058. The format of the expression value 
is log2(FPKM + 0.1) and was rounded to one digit to reduce data size. 

```{r prepare data}
library(cosinet)
# get expression data
data(exprBC)
# get sample information
data(samplesBC)

# Check the order of the samples
all(colnames(exprBC) == samplesBC$title)

# get ER status as condition
condition <- rep(NA, nrow(samplesBC))
condition[samplesBC$er_status %in% 0] <- "ER-Negative"
condition[samplesBC$er_status %in% 1] <- "ER-Positive"
condition <- factor(condition, levels = c("ER-Negative", "ER-Positive"))
# check the number of samples for each condition
table(condition)

# genes that are expressed in less than 20 samples for any condition are removed
# the data has already been transformed to log2(FPKM+0.01), so logTrans is set 
# to FALSE
exprBC <- matPreprocess(exprMat = exprBC,
                        condition = condition,
                        minSample = 20,
                        minExp = log(0 + 0.01),
                        logTrans = FALSE)
```

## Differential co-expression network

We then perform differential co-expression (DC) analysis to obtain a DC matrix. 
`Cosinet` provides a function for z-score-based DC analysis. A matrix derived 
with other methods, as long as it represents the level of differential 
co-expression of gene pairs between two conditions, can also be used for the 
downstream analysis.

```{r DC matrix}
# calculate differential co-expression matrix
dcMat <- DCMat(exprMat = exprBC,
               condition = condition,
               corMethod = "spearman")
```

## Gene centrality

Next, we rank genes based on the eigenvector centrality of the full DC network.
Eigenvector centrality measures node importance by summing its neighbors' 
centralities. A node with many connections to other high-centrality nodes has 
high eigenvector centrality, while a node with few connections has a low 
centrality value. This allows identifying highly central genes that 
participate in a larger regulatory networks of certain in biological processes 
or pathways.

```{r gene centrality, fig.width=5, fig.height=5}
# calculate gene centrality based on differential correlation matrix
geneCentrality <- eigenCentrality(mat = dcMat)
# plot gene centrality rank plot
rankPlot(geneCentrality, nLabel = 10, nIQR = 2, 
         main = "Gene Centrality")
```

## Gene set enrichment analysis

Now that we have the gene centrality scores, we can use them as input to perform 
gene set enrichment analysis (GSEA) to obtain function-specific gene sets that 
are enriched for differentially co-expressed genes. Here we use
`r Biocpkg("clusterProfiler")` and MsigDB hallmark gene sets to perform GSEA as 
an example.

```{r GSEA install, eval=FALSE}
# install packages
if (!require("clusterProfiler", quietly = TRUE)) {
    install.packages("clusterProfiler")
}
if (!require("msigdbr", quietly = TRUE)) BiocManager::install("msigdbr")
```

```{r GSEA }
# get hallmark gene sets
geneSets <- msigdbr::msigdbr(species = "Homo sapiens", category = "H")
geneSets <- geneSets[, c("gs_name", "gene_symbol")]

# run GSEA
# remember to set scoreType to "pos" since we only care about gene
# sets that are enriched at the top of the ranked list
library(clusterProfiler)
resGSEA <- GSEA(geneCentrality,
                TERM2GENE = geneSets,
                pvalueCutoff = 0.01,
                pAdjustMethod = "hochberg",
                seed = TRUE,
                scoreType = "pos",
                eps = 0)
slot(resGSEA, "result")[, c(5, 7)]
```


The third enriched gene set is "HALLMARK_ESTROGEN_RESPONSE_EARLY", we use
the core enrichment genes in this gene set to calculate the Cosinet score.

```{r GSEA get genes, fig.width=9, fig.height=4.5, fig.wide=TRUE}
# check GSEA plot for the "HALLMARK_ESTROGEN_RESPONSE_EARLY" gene set
# gseaplot(resGSEA,
#          geneSetID = slot(resGSEA, "result")$ID[3],
#          title = slot(resGSEA, "result")$Description[3])

# get core enrichment genes
genes <- slot(resGSEA, "result")$core_enrichment[3]
genes <- strsplit(genes, "\\/")[[1]]

# check DC plot for selected genes
# blue edges indicate positive correlation, red for negative correlation
# for more details, please see the manual page of the "DCPlot" function
p <- DCPlot(exprMat = exprBC,
            condition = condition,
            genes = genes,
            dcMat = dcMat,
            posDCCut = 3,
            negDCCut = -3,
            corMethod = "spearman",
            minimum = 0.3,
            labelCex = 0.6)

```

## Get Cosinet score

With a set of interesting DC genes we can calculate the corresponding Cosinet
scores based on the sub-network of these genes.

```{r getCosi, fig.width=6, fig.height=3}
# calculate Cosinet scores with given genes
# only gene pairs with a DC matrix value greater than 3 or less than -3 are 
# included in the calculation
scoreMat <- getCosi(exprMat = exprBC,
                    bin = 0.1,
                    condition = condition,
                    genes = genes,
                    dcMat = dcMat,
                    weighted = TRUE,
                    posDCCut = 3,
                    negDCCut = -3,
                    threads = 2)
scoreMat[1:5, 1:5]

# plot histogram for Cosinet scores
library(ggplot2)
data <- data.frame(Cosinet = scoreMat$Cosinet)
data <- cbind.data.frame(data, samplesBC)
ggplot(data, aes(x = Cosinet, fill = condition)) +
    geom_histogram(color="#e9ecef", alpha = 0.6,
                   position = 'identity', bins = 30) +
    guides(fill = guide_legend("ER status")) +
    geom_vline(xintercept = 0, linetype="dashed",
               color = "grey") +
    labs(x = "Cosinet score (early ER response)") +
    theme_bw()
```


## Survival analysis
Finally, we perform survival analysis using Cosinet score and age as 
covariates.

```{r survival install, eval=FALSE}
# install packages
if (!require("survival", quietly = TRUE)) install.packages("survival")
if (!require("survminer", quietly = TRUE)) install.packages("survminer")
```

```{r survival, fig.width=6, fig.height=4, fig.wide=TRUE}

library(survival)
library(survminer)

# extract ER-positive samples
dataERPos <- data[data$er_status %in% 1, ]

# fit a Cox proportional hazards regression model with Cosinet score and age
survModel <- coxph(Surv(overall_survival_days,
                        overall_survival_event) ~ Cosinet + 
                       age_at_diagnosis,
                   data = dataERPos)
summary(survModel)

# fit a survival curve for patients with different levels of Cosinet scores 
# using the Kaplan-Meier method

dataERPos$CosinetScore <- NA
dataERPos$CosinetScore[dataERPos$Cosinet <= 0] <- "Low"
dataERPos$CosinetScore[dataERPos$Cosinet > 0 & 
                           dataERPos$Cosinet <= 0.5] <- "Medium"
dataERPos$CosinetScore[dataERPos$Cosinet > 0.5] <- "High"

dataERPos$CosinetScore <- factor(dataERPos$CosinetScore, 
                                 levels = c("Low", "Medium", "High"))

fit <- survfit(Surv(overall_survival_days, 
                    overall_survival_event) ~ CosinetScore, data = dataERPos)

ggsurvplot(fit,
           data = dataERPos,
           conf.int = TRUE,         
           pval = TRUE,
           legend.labs = c("Low", "Medium", "High"),
           legend.title = "Cosinet Score",
           ggtheme = theme_bw())

```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
